Learning a functional distance between organizations with unsupervised embeddings of scientific mobility

### 2020-08-13 paper/current/main.tex
change_title
revised abstract and introduction

\begin{document} %{{{
\title{Learning a functional distance between organizations with unsupervised embeddings of scientific mobility}
\title{Unsupervised embeddings of organizations capture the latent structure of global scientific mobility} %{{{
\date{\today}
\maketitle %}}}

@@

Human mobility and migration drive major societal phenomena such as the growth and evolution of cities~\autocite{wef2017migration, curiel2018citygravity},  epidemics~\autocite{kraemer2020covid, truscott2012epidemicgravity, xia2005measlesgravity}, economies~\autocite{kaluza2010cargo, kerr2011immigration}, and innovation~\autocite{kaiser2018innovation, armano2017innovation, sugimoto2017mostimpact, petersen2018multiscale, franzoni2014advantage}.
Geographic distance and population have been considered as leading factors of human mobility~\autocite{zipf1946gravity}.
However, geographic distance is becoming less relevant in the increasingly-globalized world~\autocite{czaika2018migration, czaika2018globalisation, buchan2009globalization, altbach2004globalization} in which physical barriers are shrinking while linguistic, cultural, and historical relationships are becoming more important, which calls for alternative \textit{functional distances} that take these factors into account~\autocite{boschma2005proximity,  brown1970functional, brown1970migration, kim2018functional}.
However, existing functional distance are often arbitrary and consider only single isolated facets of mobility.
Using the case study of scientific mobility, we demonstrate the a neural embedding model can successfully encode the affiliation trajectories of scientists into a vector-space representation of organizations~\autocite{mikolov2013word2vec} from which a functional distance can be derived.
We show that the proposed functional distance reflects the multi-faced structure of scientists' mobility that is not explained by geographic distance alone, capturing relations based on culture, language, and academic prestige at multiple levels of granularity.
The vector representations of organizations themselves also reveal insights into the landscape of global scientific mobility, such as a relationship between organizations' size and the norm of its vector that is consistent across most countries. 
The ability to learn scalable, dense, and meaningful high-resolution vector representations of mobility directly from the data can open up a new avenue of studying human mobility.

@@

How far apart are two places?
The question is surprisingly hard to answer when it involves human mobility.
Geographic distance has underpinned human movements, however it is becoming less relevant in a world connected by rapid transit and global airline networks.
For instance, a person living in Australia is more likely to migrate to the United Kingdom, a far-away country with similar language and culture, than to a much closer country such as Indonesia~\autocite{pew2018migration}.
Similarly, a student in South Korea is more likely to attend a university in Canada than one in North Korea~\autocite{unesco2019students}.
Although geographic distance has been used as the most prominent basis for models of mobility, such as the Gravity~\autocite{zipf1946gravity} and Radiation~\autocite{simini2012universal} models, there have been attempts to define alternative notions of distance, or ~\textit{functional distances}~\autocite{boschma2005proximity, brown1970functional, brown1970migration, kim2018functional} from real-world data or \textit{a priori} relationships between geographic entities.
Yet, functional distances are often low-resolution, at the level of countries rather than regions, cities, or organizations, and rely on somewhat arbitrary decisions about which factors are most relevant.
Moreover, functional distances have focused on only a single facet of mobility at a time, whereas real-world mobility is multi-faceted, influenced simultaneously by geography, language, culture, history, and economic opportunity.
To capture such complexity of human mobility, new approaches that incorporate its many facets into a single high-resolution functional distance are needed.

Here, we propose using a neural embedding model that draws from real-world mobility trajectories to encode the complex landscape of global mobility into a dense, continuous, and meaningful vector-space representation, from which a \textit{functional distance} between places is derived.
Using the case of scientific mobility, we demonstrate that the functional distance captures relations based on culture, language, and academic prestige, along with geographic relations, at multiple levels of granularity.

@@

Here, scientific mobility is operationalized using a researcher's organizational affiliations derived from their publication histories indexed in a comprehensive bibliographic database (see Methods).
The most fine-grained geographical unit in this data is the organization, such as a ``university'', ``research institute'', ``business'', or ``government agency''.
As a scientist's career progresses, they move between organizations or pick up additional (simultaneous) affiliations forming \textit{affiliation trajectories} (Fig.~\ref{fig:methods:mobility_sentence}).
Thus, the trajectories encode both migration and co-affiliation---the holding of multiple simultaneous co-affiliations involving the sharing of time and capital between locations---that is typical of scientific mobility~\autocite{rodrigues2016mobility, markova2016synchronous, sugimoto2017mostimpact} (see \SI).
We use the trajectories of more than three million name-disambiguated authors who were \textit{mobile}---having more than one affiliation---between 2008 and 2019, as evidenced by their publications indexed in the Web of Science database.

@@

To assess the performance of the embedding distance relative to geographic distance, we evaluate their correlation with real-world mobility data and their performance in predicting real-world mobility flows as parameters to the gravity model~\autocite{zipf1946gravity, barbosa2018human}.
The Gravity Model is a widely used mobility model~\autocite{curiel2018citygravity, lewer2008immigrationgravity, jung2008highwaygravity, hong2016busgravity, truscott2012epidemicgravity, xia2005measlesgravity} that calculates the \textit{expected flux}, $\hat{T}_{ij}$, of scientists between organizations based on their populations and distance:

\begin{equation}
	\label{eq:gravity_basic}
	\hat{T_{ij}} = Cm_{i}m_{j}f(r_{ij}),
\end{equation}
where $m_{i}$ is the population of mobile and non-mobile authors who affiliated with institution $i$, $f(r_{ij})$ is a decay function with respect to distance between organizations, and $C$ is a constant estimated from data (see Data and Methods). 
$\hat{T}_{ij}$, which is often referred to as ``expected flux''~\autocite{simini2012universal}, is the expected frequency of the co-occurrence of organizations $i$ and $j$ in the trajectory of affiliations in the gravity model.
The gravity model dictates that the expected flow, $\hat{T}_{ij}$, is symmetric (\ie $\hat{T}_{ij }= \hat{T}_{ji})$, is proportional to the organizations' sizes, $\hat{T}_{ij} \propto m_{i}, m_{j}$, and decays as a function of their distance, $f(r_{ij})$. 
We define the distance function in terms of either the geographic distance between organizations or their embedding distance.
For geographic distance, we define $f(r_{ij})$ as the standard power-law function, and for the embedding distance, we use the exponential function, selected as the best performing for each case (Fig.~\ref{fig:supp:predict_emb_bymodel} and Fig.~\ref{fig:supp:predict_geo_bymodel}).

>>>>

\begin{document} %{{{
\title{Learning a functional distance between organizations with unsupervised embeddings of scientific mobility}
\title{Unsupervised embeddings of career trajectories capture the latent structure of global scientific mobility} %{{{
\date{\today}
\maketitle %}}}

@@

Human mobility and migration drive major societal phenomena such as the growth and evolution of cities~\autocite{wef2017migration, curiel2018citygravity},  epidemics~\autocite{kraemer2020covid, truscott2012epidemicgravity, xia2005measlesgravity}, economies~\autocite{kaluza2010cargo, kerr2011immigration}, and innovation~\autocite{kaiser2018innovation, armano2017innovation, sugimoto2017mostimpact, petersen2018multiscale, franzoni2014advantage}.
Historically, human mobility is strongly constrained by the physical separation-geographical distance~\autocite{zipf1946gravity}.
Recently, geographic distance is becoming less relevant in the increasingly-globalized world~\autocite{czaika2018migration, czaika2018globalisation, buchan2009globalization, altbach2004globalization} in which physical barriers are shrinking while linguistic, cultural, and historical relationships are becoming more important, which calls for alternative modeling framework for human mobility take these factors into account~\autocite{boschma2005proximity,  brown1970functional, brown1970migration, kim2018functional}.
Here, using a massive dataset of scientific mobility, we demonstrate a neural embedding model can successfully encode nuanced relationships between locations into a vector-space ~\autocite{mikolov2013word2vec}, providing an effective functional distance that reflects the multi-faced structure of scientists' mobility and capturing cultural, linguistic relations and even academic prestige at multiple levels of granularity as well as implicit hierarchical mobility structure.
Furthermore, vector representations of organizations themselves also reveal insights into the landscape of global scientific mobility, such as a relationship between organizations' size and the norm of its vector that is consistent across most countries. 
The ability to learn scalable, dense, and meaningful high-resolution vector representations of mobility directly derived from the data can open up a new avenue of studying human mobility.

@@

How far apart are two places?
The question is surprisingly hard to answer when it involves human mobility.
Although geographic distance has been constraining  human movements throughout history, it is becoming less relevant in a world connected by rapid transit and global airline networks.
For instance, a person living in Australia is more likely to migrate to the United Kingdom, a far-away country with similar language and culture, than to a much closer country such as Indonesia~\autocite{pew2018migration}.
Similarly, a student in South Korea is more likely to attend a university in Canada than one in North Korea~\autocite{unesco2019students}.
Although geographic distance has been used as the most prominent basis for models of mobility, such as the Gravity~\autocite{zipf1946gravity} and Radiation~\autocite{simini2012universal} models, there have been attempts to define alternative notions of distance, or ~\textit{functional distances}~\autocite{boschma2005proximity, brown1970functional, brown1970migration, kim2018functional} from real-world data or \textit{a priori} relationships between geographic entities.
Yet, functional distances are often low-resolution, at the level of countries rather than regions, cities, or organizations, and rely on somewhat arbitrary decisions about which factors are most relevant.
Moreover, functional distances have focused on only a single facet of mobility at a time, whereas real-world mobility is multi-faceted, influenced simultaneously by geography, language, culture, history, and economic opportunity.
To capture such complexity of human mobility, new approaches that incorporate its many facets into a single high-resolution functional distance are needed.

Here, we propose using a neural embedding model that draws from real-world mobility trajectories can encode the complex landscape of human mobility into a dense, continuous, and meaningful vector-space representation, from which not only a meaningful \textit{functional distance} between places but also relationship based on culture, language, and even prestige along with geographic relationship can be probed.

@@

Here, each trajectory is operationalized using a researcher's organizational affiliations derived from their publication histories indexed in a comprehensive bibliographic database (see Methods).
The most fine-grained geographical unit in this data is the organization, such as a ``university'', ``research institute'', ``business'', or ``government agency''.
As a scientist's career progresses, they move between organizations or pick up additional (simultaneous) affiliations forming \textit{affiliation trajectories} (Fig.~\ref{fig:methods:mobility_sentence}).
Thus, the trajectories encode both migration and co-affiliation---the holding of multiple simultaneous co-affiliations involving the sharing of time and capital between locations---that is typical of scientific mobility~\autocite{rodrigues2016mobility, markova2016synchronous, sugimoto2017mostimpact} (see \SI).
We use the trajectories of more than three million name-disambiguated authors who were \textit{mobile}---having more than one affiliation---between 2008 and 2019, as evidenced by their publications indexed in the Web of Science database.

@@

To assess the performance of the embedding distance relative to geographic distance, we evaluate their correlation with real-world mobility data and their performance in predicting real-world mobility flows as parameters to the gravity model~\autocite{zipf1946gravity, barbosa2018human}.
The Gravity Model is a widely used mobility model~\autocite{curiel2018citygravity, lewer2008immigrationgravity, jung2008highwaygravity, hong2016busgravity, truscott2012epidemicgravity, xia2005measlesgravity} that calculates the \textit{expected flux}, $\hat{T}_{ij}$, of peoples between locations(organizations) based on their populations and distance:

\begin{equation}
	\label{eq:gravity_basic}
	\hat{T_{ij}} = Cm_{i}m_{j}f(r_{ij}),
\end{equation}
where $m_{i}$ is the population of mobile and non-mobile authors who affiliated with institution $i$, $f(r_{ij})$ is a decay function with respect to distance between organizations, and $C$ is a constant estimated from data (see Data and Methods). 
$\hat{T}_{ij}$, which is often referred to as ``expected flux''~\autocite{simini2012universal}, is the expected frequency of the co-occurrence of organizations $i$ and $j$ in the trajectory of affiliations in the gravity model.
The gravity model dictates that the expected flow, $\hat{T}_{ij}$, is symmetric (\ie $\hat{T}_{ij }= \hat{T}_{ji})$, is proportional to the organizations' sizes, $\hat{T}_{ij} \propto m_{i} m_{j}$, and decays as a function of their distance, $f(r_{ij})$. 
We define the distance function in terms of either the geographic distance between organizations or their embedding distance.
For geographic distance, we define $f(r_{ij})$ as the standard power-law function, and for the embedding distance, we use the exponential function, selected as the best performing for each case (Fig.~\ref{fig:supp:predict_emb_bymodel} and Fig.~\ref{fig:supp:predict_geo_bymodel}).
